Activity Monitoring Device Data Project
========================================================

Utilizing data from a personal activity monitoring device, this analysis attempts to answer several questions about patterns in the number of steps taken by an individual over the course of 2 months; October and Novermber 2012.

Load all necessary packages here.
```{r packages, results='HIDE'}
library(Hmisc)
library(lattice)
library(plyr)
```

## Loading and Preprocessing the Data
We begin by loading and reading the data and changing the dates into workable dates.

```{r load}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(url, destfile=("./newfile.zip"), method="curl")
unzip("newfile.zip")
data <- read.csv("activity.csv")
dates <- data$date
dates <- as.Date(as.character(dates), "%Y-%m-%d")
data$date <- dates
```

At this point, the data is sufficiently ready for analysis and to help us answer our questions. 

## What is mean total number of steps taken per day?
First, use the tapply function to plot a histogram of total number of steps each day and we see a pretty normal distribution.

```{r histogram}
hist(tapply(data$steps, data$date, FUN=sum), breaks=10, xlab="# of Steps", main="Hist of Total # of Steps per Day")
```

Next we find the mean and median of total steps each day which come out to be about 9,354 and 10,765 respectively.

```{r tendency measurements}
meanSteps <- sum(data$steps, na.rm=TRUE)/length(unique(data$date))
meanSteps
medSteps <- (aggregate(steps~date, data=data, FUN=sum))
median(medSteps$steps)
```

## What is the average daily activity pattern?
We use a line chart showing the average # of steps in each time interval to discover the 5-minute interval that has the highest average steps is a little bit before 2PM.

```{r}
intervalData <- aggregate(steps~interval, data=data, FUN=mean)
plot(intervalData, type="l", xlab="5-minute Interval", ylab="Number of Steps", main="Average # of Steps per 5-minute Interval")
findMax <- subset(intervalData, intervalData$steps==max(intervalData$steps))
findMax
```

## Imputing missing values
Since the rate of occurance of NA values is about 13% --- quite high --- we use a simple mean method to impute these missing values. We can see from the boxplot below that this system spreads teh data out a little more since the mean is about 10,000 and we have a lot of zero values. This is not ideal and granted more time, I would consider using a better imputation package such as VIM. 

```{r}
sum(is.na(data))/nrow(data)
data$imputed_steps <- with(data, impute(steps, mean))
hist(tapply(data$imputed_steps, data$date, FUN=sum), breaks=10, xlab="# of Steps", main="Hist of Total # of Steps per Day with Imputed Value")
impMean <- aggregate(imputed_steps~date, data=data, FUN=mean)
impMedian <- aggregate(imputed_steps~date, data=data, FUN=median)
boxplot(data$steps, data$imputed_steps)
```

## Are there differences in activity patterns between weekdays and weekends?
Using the lattice and plyr packages, we convert dates into weekends/weekdays to show that this individual took more steps on weekdays in November and took more steps on the weekends in October.

```{r weekdays and weekends}
data$weekend <- weekdays(data$date)=="Saturday" | weekdays(data$date)=="Sunday"
data$weekend <- as.factor(data$weekend)
data$weekend <- revalue(data$weekend, c("TRUE"="weekday", "FALSE"="weekend"))
xyplot(imputed_steps~interval | weekend, data, type="l", layout=c(1,2))
```
